### 2019-09-14 ic-journal/multivariate.tex
Merged changes in multivariate.tex (with minor fixes)

The central problem of the \multivariate formalisation is the
representation of CCS equations (expressions, contexts).
\hlD{In the} univariate formalisation, $\lambda$-functions (of type
$CCS\rightarrow CCS$) are used to represent univariate CCS
\hlD{equations, and variable substitutions are simply 
calls to $\lambda$-functions on the target CCS terms.}
This idea cannot be extended to the \multivariate case, as
we do not have a fixed number of variables to deal with.

@@

\finish{too many changes below to highlight them}

@@

We denote the set of all bound variables of a
given CCS term $E$ (we recall that these are variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'', i.e. functions taking CCS terms returning sets of
variables (of type \HOLinline{\ensuremath{\alpha}}).
% Milner didn't give their definitions explicitly. However, one can
% imagine that, for most CCS operators they just pass into
% subterms without collecting anything (and $\nil$ gives $\emptyset$).
For their definition the interesting cases are those of 
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
it does not  contains free variables, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLSymConst{\HOLTokenDefEquality{}} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}

Note that, in CCS expression, the set of free and bound
variables need not be disjoint, e.g.\ in  $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables may
decrease. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
difficulities when doing proofs about CCS transitions. As an evidence,
we prove the following fundamental result about $\fv{\cdot}$:

@@

lemma, and obtain \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E}} 
from the transitivity of $\subseteq$:

@@

Knowing \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E}} we cannot prove \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E} \HOLConst{DELETE} \HOLFreeVar{X}}. However, if we knew \HOLinline{\HOLFreeVar{X} \HOLSymConst{\HOLTokenNotIn{}} \HOLConst{FV} \HOLFreeVar{E\sp{\prime}}}, then \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLConst{DELETE} \HOLFreeVar{X} \HOLSymConst{\ensuremath{=}} \HOLConst{FV} \HOLFreeVar{E\sp{\prime}}}, and 

@@

   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}}
     \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \ensuremath{(}\HOLFreeVar{fm} \HOLSymConst{\ensuremath{\setminus}} \HOLFreeVar{X}\ensuremath{)} \HOLFreeVar{E}\ensuremath{)}
     \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \HOLFreeVar{fm} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_SUBST_rec]}
\end{alltt}
Note that the variable substitution only occurs on free
variables. In the case of $\recu X E$: if $X$ (a bound variable) is in
the domain of \HOLinline{\HOLFreeVar{fm}}, \HOLinline{\HOLConst{CCS_SUBST}} must continue the
substitution on $E$ using a reduced map~--- without $X$~--- so that all
occurences of $X$ in $E$ are correctly bypassed. Since
 since \HOLinline{\HOLConst{CCS_SUBST}} only runs
once on each  subterm, the possible free variables in  \HOLinline{\HOLFreeVar{Ps}} are never
 substituted. 
% Thus the process is indeed
% order-independent. 
Below we present some key lemmas about
\HOLinline{\HOLConst{CCS_SUBST}}, omitting the  proofs:

\begin{lemma}[\texttt{CCS_SUBST_elim}]
If the set of  free variables of $E$ is disjoint from \HOLinline{\HOLFreeVar{Xs}}, a
substitution of \HOLinline{\HOLFreeVar{Xs}} in $E$ does not change $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{E}
\end{alltt}
\end{lemma}

\begin{lemma}[\texttt{CCS_SUBST_self}]
Substituting each free variable $X$ in $E$ to \HOLinline{\HOLConst{var}\\\;\HOLFreeVar{X}} (itself)
does not change $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLConst{MAP} \HOLConst{var} \HOLFreeVar{Xs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{E}
\end{alltt}
\end{lemma}

The next lemma plays an important role. It essentially swaps the order of two substitutions:
\begin{lemma}[\texttt{CCS_SUBST_nested}]
Under certain conditions, two nested substitutions can be
converted into a single substitution where the targets are substituted first:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{C}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}
\end{alltt}
\end{lemma}

The next three lemmas show the relative correctness of \texttt{CCS\_SUBST}
w.r.t. \texttt{CCS\_Subst}:
\begin{lemma}[\texttt{CCS_SUBST_SING}]
  If there is only one single variable $X$ in the map (with the target
  expression $E'$), then \HOLinline{\HOLConst{CCS_SUBST}} behaves exactly the
  same as (the univariate) \HOLinline{\HOLConst{CCS_Subst}}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_SUBST} \ensuremath{(}\HOLConst{FEMPTY} \HOLSymConst{|+} \ensuremath{(}\HOLFreeVar{X}\HOLSymConst{,}\HOLFreeVar{E\sp{\prime}}\ensuremath{)}\ensuremath{)} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLConst{CCS_Subst} \HOLFreeVar{E} \HOLFreeVar{E\sp{\prime}} \HOLFreeVar{X}
\end{alltt}
\end{lemma}

@@

Finally, the following two lemmas precisely estimate the free and
bound variables of a substituted term:
\begin{lemma}[\texttt{BV_SUBSET_BIGUNION}]
The bound variables of \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}} contains
at most the union of all bound variables from $E$ and \HOLinline{\HOLFreeVar{Ps}}.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLFreeVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BIGUNION} \ensuremath{(}\HOLConst{IMAGE} \HOLConst{BV} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Ps}\ensuremath{)}\ensuremath{)}
\end{alltt}
\end{lemma}

\begin{lemma}[\texttt{FV_SUBSET_BIGUNION_PRO}]
The free variables of \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}} contain
at most the union of all free variables from $E$ and \HOLinline{\HOLFreeVar{Ps}},
excluding \HOLinline{\HOLFreeVar{Xs}}.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FV} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E} \HOLConst{DIFF} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BIGUNION} \ensuremath{(}\HOLConst{IMAGE} \HOLConst{FV} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Ps}\ensuremath{)}\ensuremath{)}
\end{alltt}
\end{lemma}

@@

\subsection{\Multivariate (weakly-guarded) contexts}

>>>>

The central problem of the \multivariate formalisation is the
representation of CCS equations (expressions, contexts).
\hlD{In the} \univariate formalisation, $\lambda$-functions (of type
$CCS\rightarrow CCS$) are used to represent univariate CCS
\hlD{equations, and variable substitutions are simply 
calls to $\lambda$-functions on the target CCS terms.}
This idea cannot be extended to the \multivariate case, as
we do not have a fixed number of variables to deal with.

@@

@@

We denote the set of all bound variables of a
given CCS term $E$ (we recall that these are variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'', i.e. functions taking CCS terms returning sets of
variables (of type \HOLinline{\ensuremath{\alpha}}).
% Milner didn't give their definitions explicitly. However, one can
% imagine that, for most CCS operators they just pass into
% subterms without collecting anything (and $\nil$ gives $\emptyset$).
For their definition the interesting cases are those of 
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
\hl{it does not contain any free variable}, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLSymConst{\HOLTokenDefEquality{}} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}
And a list of CCS processes can be asserted by \HOLinline{\HOLConst{ALL_PROC}} defined upon \HOLinline{\HOLConst{IS_PROC}}:
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{ALL_PROC} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenEquiv{}} \HOLConst{EVERY} \HOLConst{IS_PROC} \HOLFreeVar{Es}\hfill{[ALL_PROC_def]}
\end{alltt}

Note that, in CCS expression, the set of free and bound
variables need not be disjoint, e.g.\ in  $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables \hl{either
remains the same or decreases}. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
difficulities when doing proofs about CCS transitions. As an evidence,
we prove the following fundamental result about $\fv{\cdot}$:

@@

lemma, and obtain \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E}} by the transitivity of $\subseteq$:

@@

Knowing \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E}} we cannot prove \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E} \HOLConst{DELETE} \HOLFreeVar{X}}. However, if we knew \HOLinline{\HOLFreeVar{X} \HOLSymConst{\HOLTokenNotIn{}} \HOLConst{FV} \HOLFreeVar{E\sp{\prime}}}, then \HOLinline{\HOLConst{FV} \HOLFreeVar{E\sp{\prime}} \HOLConst{DELETE} \HOLFreeVar{X} \HOLSymConst{\ensuremath{=}} \HOLConst{FV} \HOLFreeVar{E\sp{\prime}}}, and \hl{then}

@@

   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenDefEquality{}}
     \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \ensuremath{(}\HOLFreeVar{fm} \HOLSymConst{\ensuremath{\setminus}} \HOLFreeVar{X}\ensuremath{)} \HOLFreeVar{E}\ensuremath{)}
     \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \HOLFreeVar{fm} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_SUBST_rec]}
\end{alltt}
Note that the variable substitution only occurs on free
variables. In the case of $\recu X E$: if $X$ (a bound variable) is in
the domain of \HOLinline{\HOLFreeVar{fm}}, \HOLinline{\HOLConst{CCS_SUBST}} must continue the
substitution on $E$ using a reduced map~--- without $X$~--- so that all
occurences of $X$ in $E$ are correctly bypassed.
\hl{Since} \HOLinline{\HOLConst{CCS_SUBST}} only runs
once on each subterm, the possible free variables in  \HOLinline{\HOLFreeVar{Ps}} are never
 substituted. 
% Thus the process is indeed
% order-independent. 
Below we present some key lemmas about
\HOLinline{\HOLConst{CCS_SUBST}}, omitting the proofs:\footnote{Hereafter,
  some new logical constants from HOL's \texttt{pred_set}
and \texttt{list} theories are used (c.f.~\cite{holdesc} for more
details.): \HOLinline{\HOLConst{DISJOINT}} denotes set disjointness;
``\HOLinline{\HOLConst{set}\\\;\HOLFreeVar{Xs}}'' is the set converted from the list \HOLinline{\HOLFreeVar{Xs}};
\HOLinline{\HOLConst{ALL_DISTINCT}} says
all elements of a list are distinct with each other; \HOLinline{\HOLConst{MAP}} is
the mapping function from one list to another; \HOLinline{\HOLConst{EVERY}} means
each element of a list satisfies a predicate; \HOLinline{\HOLConst{ZIP}} creates a
new list from two lists of the same length, and each element of the new list is a pair of
elements, each from one of the old lists.}

\begin{lemma}[\texttt{CCS_SUBST_elim}]
If the free variables of $E$ is disjoint with \HOLinline{\HOLFreeVar{Xs}}, a
substitution of \HOLinline{\HOLFreeVar{Xs}} in $E$ does not change $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{E}
\end{alltt}
\end{lemma}

\begin{lemma}[\texttt{CCS_SUBST_self}]
  Substituting each free variable $X$ % of \HOLinline{\HOLFreeVar{Xs}}
  in $E$ to \HOLinline{\HOLConst{var}\\\;\HOLFreeVar{X}} (itself) does not change $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLConst{MAP} \HOLConst{var} \HOLFreeVar{Xs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{E}
\end{alltt}
\end{lemma}

The next lemma plays an important role. It essentially swaps the order of two substitutions:
\begin{lemma}[\texttt{CCS_SUBST_nested}]
Under certain conditions \hl{(to get rid of substitution orders)}, two nested substitutions can be
converted into a single substitution where the targets are substituted first:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Es} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}}
   \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{C}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{C}
\end{alltt}
\end{lemma}

The next three lemmas show the relative correctness of \texttt{CCS\_SUBST}
w.r.t.~\texttt{CCS\_Subst}:
\begin{lemma}[\texttt{CCS_SUBST_sing}]
  If there is only one single variable $X$ in the map (with the target
  expression $E'$), then \HOLinline{\HOLConst{CCS_SUBST}} behaves exactly the
  same as (the univariate) \HOLinline{\HOLConst{CCS_Subst}}:
\begin{alltt}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{[}\HOLFreeVar{E\sp{\prime}}\ensuremath{]}\ensuremath{/}\ensuremath{[}\HOLFreeVar{X}\ensuremath{]}\ensuremath{]} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\HOLFreeVar{E\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}
\end{alltt}
\end{lemma}

@@

Finally, the following two lemmas precisely estimate the free and
bound variables of a substituted term:
\begin{lemma}[\texttt{BV_SUBSET_BIGUNION}]
The bound variables of \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}} have
at most the union of all bound variables from $E$ and \HOLinline{\HOLFreeVar{Ps}}.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLFreeVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BIGUNION} \ensuremath{(}\HOLConst{IMAGE} \HOLConst{BV} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Ps}\ensuremath{)}\ensuremath{)}
\end{alltt}
\end{lemma}

\begin{lemma}[\texttt{FV_SUBSET_BIGUNION_PRO}]
The free variables of \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}} have
at most the union of all free variables from $E$ and \HOLinline{\HOLFreeVar{Ps}},
excluding \HOLinline{\HOLFreeVar{Xs}}.
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{ALL_DISTINCT} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{LENGTH} \HOLFreeVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FV} \ensuremath{(}\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLFreeVar{E} \HOLConst{DIFF} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BIGUNION} \ensuremath{(}\HOLConst{IMAGE} \HOLConst{FV} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Ps}\ensuremath{)}\ensuremath{)}
\end{alltt}
\end{lemma}

@@

\subsection{\Multivariate (weakly guarded) contexts}

Recall in Section~\ref{ss:context}, univariate contexts and their
guarded companions are defined as predicates over $\lambda$-expressions
of type $CCS\rightarrow CCS$. These predicates, such as
\HOLinline{\HOLConst{CONTEXT}}
(contexts/expressions), \HOLinline{\HOLConst{WG}} (weak guarded contexts),
\HOLinline{\HOLConst{SG}} (strongly guarded contexts) and \HOLinline{\HOLConst{SEQ}} (sequential
contexts), are all defined inductively, i.e.~they are built from some ``holes''
in a bottom-up way. For instance, \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{\mid}}\\\;\HOLFreeVar{P}\ensuremath{)}}
holds because \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}}
holds \hl{as a base case of the inductive definition of} \HOLinline{\HOLConst{WG}}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{a}. \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}\hfill{[WG1]}
\end{alltt}
\hl{For} any agent variable $X$, we have \hl{by $\beta$-reduction:}
\HOLinline{\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{var}\\\;\HOLFreeVar{X}\ensuremath{)}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}} (or ``$l.X + P$'' in textbook
notation) is weakly guarded (c.f.~Def.~\ref{def:guardness}).
